Improving \textsl{Gaia} parallaxes with a data-driven model of the color--magnitude diagram

### 2017-03-14 paperDistance/paper.tex
bibliography

%@archiver{PUT FIGURE NAME HERE}

% # To-do list
% - We need a standard name for the \tgas--\tmass--\PanSTARRS\ Intersection.
% - zeroth draft the introduction
% - zeroth draft the method section
% - zeroth draft the assumptions list
% - reformat single-column

@@

\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tgas}{\project{\acronym{TGAS}}}
\newcommand{\tmass}{\project{\acronym{2MASS}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\xd}{\acronym{XD}}
\newcommand{\cmd}{\acronym{CMD}}

@@

The novel use of \xd\ here is to treat its output---the deconvolved \cmd---as
a prior for use in inference of individual-star distances.
These inferences, one per star, provide much more precise parallax, distance,
or absolute magnitude estimates than we get from each star's primary
\tgas\ parallax measurement alone.
That is, we are using the \xd\ model to de-noise the \gaia\ parallax
measurements.
Technically, since \xd\ produces a maximum-marginalized-likelihood estimator
of the deconvolved distribution,
its use as a prior is only approximately valid; it constitutes using the
data twice.
However, since the data set is large, the \xd\ approximation is not bad; it is
sometimes known as the ``empirical Bayes'' method, and is well studied (DWH REFS).

It wouldn't be using the data twice---that is, it would be valid from a
probabilistic perspective---if we performed a full hierarchical Bayesian
inference.
Exploration of that possibility, and its computational tractability,
is among our long-term goals and motivations.
Along those lines, this paper can be seen as a companion to
a related project (CITE BORIS), in which we use a
slightly less appropriate model for the \cmd, but in which we take a much more principled
approach to the inference.

@@

\section{Why does this work?}

LA+DWH: Do we we want to put a pedagogical section in right here?

Hogg would propose a straight-line fit (possibly with intrinsic
width); showing that it can be used to de-noise noisy y-data.

@@

\begin{itemize}
\item Some kind of stationarity assumption.
\item Our empirical bayes approach is ok---that is, our numbers are large enough.
\item The photometric uncertainties in the J band are small compared with the uncertainties in the parallax; the uncertainties are gaussian
\item number of gaussians
\item no use of physical models, baby!
\end{itemize}

@@

\begin{equation}
\label{eq:data}
\begin{aligned}
\mathrm{data vector} \;\;\;\; &\vec{y} = [10^{0.2M_{J,n}}, (J- K)_n] \;  n \in \mathrm{stars} \\
\mathrm{with \; absolute \; magnitude} \;\;\;\; &M_{J,n} = J_n - DM_n - Q_JE(B-V)_n \\
& DM_n = -5\log_{10}\frac{1}{\varpi} \\
\mathrm{and \; color} \;\;\;\; &(J - K)_n = J_n - K_n - Q_{JK}E(B-V)_n
\end{aligned}
\end{equation}

where $E(B-V)$ is the output of the Schlafly dust model, and $Q$ is the correction factor for that band assuming some reddening law. To generate our prior, we fit the data in the CMD space using XD (Bovy et al) which deconvolves the data with its noise, which are heterogeneous measurement uncertainties. It generates the denoised, underlying distribution from which the noisey data were drawn from. It represents the distribution as a mixture of gaussians which maximizes the likelihood of the noisy data.

@@

%\begin{thebibliography}{}
%\end{thebibliography}

>>>>

% # To-do list
% - We need a standard name for the \tgas--\tmass\ Intersection.
% - zeroth draft the introduction
% - zeroth draft the method section
% - zeroth draft the assumptions list
% - reformat single-column

@@

\newcommand{\bibtex}{\textsc{Bib}\!\TeX} % bibtex. Not quite the correct typesetting, but close enough
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tgas}{\project{\acronym{TGAS}}}
\newcommand{\tmass}{\project{\acronym{2MASS}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\xd}{\acronym{XD}}
\newcommand{\cmd}{\acronym{CMD}}

@@

The novel use of \xd\ \citep{bovy11} here is to treat its output---the deconvolved \cmd---as
a prior for use in inference of individual-star distances.
These inferences, one per star, provide much more precise parallax, distance,
or absolute magnitude estimates than we get from each star's primary
\tgas\ parallax measurement alone.
That is, we are using the \xd\ model to de-noise the \gaia\ parallax
measurements.
Technically, since \xd\ produces a maximum-marginalized-likelihood estimator
of the deconvolved distribution,
its use as a prior is only approximately valid; it constitutes using the
data twice.
However, since the data set is large, the \xd\ approximation is not bad; it is
sometimes known as the ``empirical Bayes'' method, and is well studied (DWH REFS).

It wouldn't be using the data twice---that is, it would be valid from a
probabilistic perspective---if we performed a full hierarchical Bayesian
inference.
Exploration of that possibility, and its computational tractability,
is among our long-term goals and motivations.
Along those lines, this paper can be seen as a companion to
a related project (CITE BORIS), in which we use a
somewhat less flexible model for the \cmd, but in which we take a much more principled
approach to the inference.

@@

@@

\begin{itemize}
\item The photometric uncertainties in the J band are small compared with the uncertainties in the parallax
\item Our empirical bayes approach is ok as long as we only infer information about the data contained within the prior
\item the uncertainties are gaussian
\item most stars lie where we expect them to on the CMD
\item number of gaussians that's efficient? or do x-validation
\item can't think
\end{itemize}

@@

\begin{equation}
\label{eq:data}
\begin{aligned}
\mathrm{data vector} \;\;\;\; &\vec{y} = [10^{0.2M_{J,n}}, (J- K)_n] \;  n \in \mathrm{stars} \\
\mathrm{with \; absolute \; magnitude} \;\;\;\; &M_{J,n} = J_n - DM_n - Q_JE(B-V)_n \\
& DM_n = -5\mathrm{log_{10}}\frac{1}{\varpi} \\
\mathrm{and \; color} \;\;\;\; &(J - K)_n = J_n - K_n - Q_{JK}E(B-V)_n

\end{aligned}
\end{equation}

where $E(B-V)$ is the output of the Schlafly dust model, and $Q$ is the correction factor for that band assuming some reddening law. To generate our prior, we fit the data in the CMD space using XD which deconvolves the data with its noise, which are heterogeneous measurement uncertainties. It generates the denoised, underlying distribution from which the noisey data were drawn from. It represents the distribution as a mixture of gaussians which maximizes the likelihood of the noisy data.

@@

\bibliography{lmanders}
%\begin{thebibliography}{}
%\end{thebibliography}
